---
title: "Data Preparation for 2020-2024 Preterm Births in US Territories Report [COMPLETE EXAMPLE]"
author: "K Labgold"
date: "`r Sys.Date()`"
output:
  word_document: 
    reference_docx: "mod_reference_template.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(rio, dplyr) # we list the packages we need here
```

## 0. Overview

Our objective is to create a report on preterm births for US territories during 2020-2024 using the publicly available live births data from XXXX.

This document will cover the R syntax for your standard operations of:
    + Load Data
    + Explore data using basic summaries
    + Join datasets together based on a shared column id
    + Row bind datasets together
    + Subset to only necessary columns
    + Filter to specific conditions
    + Create variables of different types (i.e., character, numeric, factor, etc.)
    + Create new variables from existing variables

## 1. Load Initial Datasets

In this example we'll be loading in two excel files, each with two sheets. Let's use the `rio` package to pull in the 2020-2023 excel file.

We will need to pull in each sheet separately: `demographic` and `clinical`

```{r}
demographic.18to23 <- rio::import("training_data_files/territory_births_18to23.xlsx", # excel file path
                           which = "demographic") # sheet name

clinical.18to23 <- rio::import("training_data_files/territory_births_18to23.xlsx",  # excel file path
                           which = "clinical") # sheet name
```

We should now see two objects for the datasets we loaded in our Environment window. In the environment window we can also see the number of observations and the number of variables in each object.

**Your turn!**

Create two objects, one named `demographic.24` that pulls in the 2024 births data demographic variables sheet, and `clinical.24` that pulls in the 2024 births data clinical variables sheet.

```{r}
demographic.24 <- rio::import("training_data_files/territory_births_24.xlsx", which = "demographic")

clinical.24 <- rio::import("training_data_files/territory_births_24.xlsx", which = "clinical")
```


For general awareness, R can load many other data types including: 
    + XXX
    + XXX
    + XXX

**Your turn!**

Take a minute to use the help function to explore package functionality for 

```{r}

```

## 2. Explore Data

We can get a quick view of our data in the environment window by clicking the arrow next to the object name.

If you click on the data, it will open as a new tab to view. This operates the same as calling the function:

```{r}
View(demographic.24)
```

names()

Summary() function

Class type

table() function

count missings



## 3. Join Data


## 4. Rowbind Data


## 5. 


