---
title: "Data Preparation for 2020-2024 Preterm Births in US Territories Report" 
subtitle: "[COMPLETE EXAMPLE]"
author: "K Labgold"
date: "`r Sys.Date()`"
output:
  word_document: 
    reference_docx: "mod_reference_template_v2.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(rio, dplyr) # we list the packages we need here
```

# Overview

Our objective is to create a report on preterm births for US territories during 2020-2024 using the publicly available live births data from XXXX.

This document is split into X parts.

# Part 1. Preparing our analytic data set

Part 1 will cover the R syntax for your standard operations of: 

- Load Data 
- Explore data using basic summaries 
- Join datasets together based on a shared column id 
- Row bind datasets together 
- Subset to only necessary columns 
- Filter to specific conditions 
- Create variables of different types (i.e., character, numeric, factor, etc.) 
- Create new variables from existing variables 

## 1. Load Initial Datasets

In this example we'll be loading in two excel files, each with two sheets. Let's use the `rio` package to pull in the 2020-2023 excel file.

We will need to pull in each sheet separately: `demographic` and `clinical`

```{r}
demographic.18to23 <- rio::import("training_data_files/territory_births_18to23.xlsx", # excel file path
                           which = "demographic") # sheet name

clinical.18to23 <- rio::import("training_data_files/territory_births_18to23.xlsx",  # excel file path
                           which = "clinical") # sheet name
```

We should now see two objects for the datasets we loaded in our Environment window. In the environment window we can also see the number of observations and the number of variables in each object.

### Try it on your own!

Create two objects, one named `demographic.24` that pulls in the 2024 births data demographic variables sheet, and `clinical.24` that pulls in the 2024 births data clinical variables sheet.

```{r}
demographic.24 <- rio::import("training_data_files/territory_births_24.xlsx", which = "demographic")

clinical.24 <- rio::import("training_data_files/territory_births_24.xlsx", which = "clinical")
```


For general awareness, R can load many other data types including: 

- XXX
- XXX
- XXX

### Try it on your own!

Take a minute to first read about the haven package: https://haven.tidyverse.org/

Then use the help function below to explore the function that loads data from your typical statistical analysis software package (e.g., SAS, SPSS, STATA).

```{r, eval = FALSE}
# SAS
?haven::read_sas()

# SPSS
?haven::read_spss()

# STATA
?haven::read_stata()
```

## 2. Explore Data

Let's start with exploring demographic.18to23 file. We can get a quick view of our data in the environment window by clicking the arrow next to the object name.

If you click on the data, it will open as a new tab. This operates the same as calling the function:

```{r}
View(demographic.18to23) # opens data frame as a new tab
```

Let's explore further to get:
    + Names of the columns in our data frame using `names()`
    + Basic attributes of each variable using `str()` and `summary()`
    + Explore first few or last few rows with `head()` and `tail()`
    + Class type of a variable `class()`
    + Get a table of variable values `table()` *Note this function can also be used for crosstabs!*
    + Count the number of missing values using `sum()` with `is.na()`
    
    
```{r}
####----Demographic 18 to 23----####

# Columns in this data frame
names(demographic.18to23)

# Structure
str(demographic.18to23)

# First few or last few rows
head(demographic.18to23)
tail(demographic.18to23)

# Basic attributes of each variable
summary(demographic.18to23)

# Class type of a variable
class(demographic.18to23$DOB_YY)

# Table of variable
table(demographic.18to23$DOB_YY, useNA = "always") # useNA = "always tells the table to include a count for any NA-type missings

# Crosstab
table("Year" = demographic.18to23$DOB_YY, 
      "Race & Ethnicity" = demographic.18to23$MRACEHISP, 
      useNA = "always")

# Can also count missing values using is.na() nested in sum()
sum(is.na(demographic.18to23$DOB_YY))
```


### Try it on your own!

Explore the remaining three objects to get a sense of data structure.

```{r}
####----Clinical 18 to 23----####

####----Demographic 2024----####

####----Clinical 2024----####

```



## 3. Join Data

For our desired output, we need to join the demographic and clinical variables for each year set. We will perform a left join using the demographic data frame as the base and joining the clinical data to it based on the shared "id". 

You can read about other join types in R here: https://dplyr.tidyverse.org/reference/mutate-joins.html

```{r}
births18to23 <- demographic.18to23 %>% # x
                  left_join(clinical.18to23, # y
                            by = c("id")) # shared id to join on 
# if the id name is not the same across data frames, we would instead use `by = c("id.x" = "id.y") 

```

### Try it on your own!

Create a new object named `births24` using a left_join from the demographic and clinical 2024 birth files.

```{r}
births24 <- demographic.24 %>%
              left_join(clinical.24,
                        by = c("id" = "id2"))
```

Using some of the explore data functions we previously used, check to confirm the join worked as intended.

```{r}
# columns in new dataset
names(births24)

# confirm values in first few rows
head(births24)



```



## 4. Rowbind Data

Now to have one working data frame, we need to rowbind the 2018-2023 and 2024 data frames. As with most tasks in R, there are many functions available to do this. I tend to recommend the `bind_rows()` function from the `dplyr` package, since it (A) will automatically align columns -- i.e. you don't have to have them all in the same order already for each data frame -- and (B) it will fill in missing columns with the proper missing variable type.

### Try it on your own!

Using the R help function for `dplyr::bind_rows()` to row bind `births18to23` and `births24`. Name the new object `births18to24`. After making the new dataframe, explore the dataframe to ensure the operation worked as intended.

```{r, eval = FALSE}
# R helper
?dplyr::bind_rows
```

```{r}
# row bind using dplyr::bind_rows()
births18to24 <- bind_rows(births18to23, births24)

# explore rowbind
View(births18to24)
table(births18to24$DOB_YY, useNA = "always")
```

**Clean out Environment**
Before we move on, let's clean up our environment a bit. We are are now working with `births18to24`, but we have several other objects still in our environment. We can remove no longer needed objects with the rm() function.

```{r}
rm(births18to23, births24, clinical.18to23, clinical.24, demographic.18to23, demographic.24)
```

If you look in your environment panel of the R Studio, you should now only have our 1 object remaining!


## 5. Filtering Rows

We have two primary filtering tasks for our report:

1. Our report is going to focus only on 2020-2024 data so we need to remove rows from 2018
2. The focus of our report is going to be on preterm births, so for this example we will need to filter to observations with gestational age information available. 

Starting with task #1: filter to 2020-2024 years only:
```{r}
class(births18to24$DOB_YY) # character class type - will need quotes!

births20to24 <- births18to24 %>%
                            filter(DOB_YY %in% c("2020", "2021", "2022", "2023", "2024"))

```

A couple things to note about the above. In R, we use the logical `==` for equivalence but since we are filter from a list, we need to use the operator `%in%`. 

You'll also notice that the years in a list `c()` and are quoted. Quotes are needed because the year class type is a character.

### Try it on your own!

What would it look like to filter only to 2020? Create a dummy dataframe named `births2020` that is only 2020 births.

```{r}
# Option 1: using ==
births2020 <- births18to24 %>%
                filter(DOB_YY == "2020")

# Option 2: using %in%
births2020 <- births18to24 %>%
                filter(DOB_YY %in% c("2020")) # Can still use %in% for a list of 1!
```


### Try it on your own!

Perform task #2: filter to births with gestational age data available [`OEGest_R3` value of 1 or 2]. Make the new object named `births18to24_step5`.

**Hint:** The negative of `==` is `!=`.

```{r}
class(births20to24$OEGest_R3) # numeric - will NOT need quotes!

births20to24_step5 <- births20to24 %>%
                        filter(OEGest_R3 != 3) 
                          
table(births20to24_step5$OEGest_R3, useNA = "always")                             


```

As a bonus, we could filter using the opposite of `%in%`. Let's use the dummy `births2020` dataset as an example.

We can create this using the built-in R function `Negate()`. Just one more great example of how flexible R is! 

```{r}
`%notin%` <- Negate(`%in%`)

births2020 <- births2020 %>%
                filter(OEGest_R3 %notin% c(3))

table(births2020$OEGest_R3, useNA = "always")              
```


Let's now remove `births2020` from the environment because it was just for practice!

```{r}
rm(births2020)
```

## 6. Coding Missing Values

`MRACEHISP` and `DMETH_REC` both have one value that is unknown/missing. 

| **MRACEHISP** | **Label**             |
| :---          | :---                  | 
| 1             | Non-Hispanic White    |
| 2             | Non-Hispanic Black    |
| 3             | Non-Hispanic American Indian and Alaska Native    |
| 4             | Non-Hispanic Asian    |
| 5             | Non-Hispanic Native Hawaiian and Other Pacific Islander   |
| 6             | Non-Hispanic Multiple Races   |
| 7             | Hispanic              |
| 8             | Origin unknown or not stated   |

| **DMETH_REC** | **Label**             |
| :---          | :---                  | 
| 1             | Vaginal               |
| 2             | C-Section             |
| 9             | Unknown               |


In R, that missing needs to be coded as a specific missing value that is consistent with it's class type. This might be:
* `NA_character_` for character class type
* `NA_integer_` for integer class type
* `NA_real_` for numeric class type
* `NA` for generic NA use

Remember, in R everything needs to be assigned because it is an object-oriented language. To do this, we will make a new variable for each of the above using the appropriate class function. 

Let's start with `MRACEHISP`, and use the `mutate()` function from the `dplyr` package to create a new variable named `MRACEHISP.na`.

We will use the `ifelse()` function to change the missing value to the appropriate class NA value. 

```{r}
# check the class of MRACEHISP
class(births20to24_step5$MRACEHISP)

# make new variable with NA_character_
births20to24_step6 <- births20to24_step5 %>%
                        mutate(MRACEHISP.na = ifelse(MRACEHISP == 8, NA_character_, MRACEHISP))
# If MRACEHISP == 8, code as NA, otherwise keep the value of MRACEHISP

### CODE CHECKS!
# check the coding worked
table(births20to24_step6$MRACEHISP.na, useNA = "always")

# cross tab
table(mracehisp = births20to24_step6$MRACEHISP,
      mracehisp.na = births20to24_step6$MRACEHISP.na,
      useNA = "always")
```


### Try it on your own!

Create a new object `births20to24_step6.pt2` from `births20to24_step6` to create a new variable `DMETH_REC.na` using the `mutate()` and `ifelse()` functions and the appropriate NA class.

```{r}
# check the class of DMETH_REC
class(births20to24_step6$DMETH_REC)

# make new variable with NA_real_
births20to24_step6.pt2 <- births20to24_step6 %>%
                            mutate(DMETH_REC.na = ifelse(DMETH_REC == 9, 
                                                         NA_character_, 
                                                         DMETH_REC))
# If DMETH_REC == 9, code as NA, otherwise keep the value of DMETH_REC

### CODE CHECKS!
# check the coding worked
table(births20to24_step6.pt2$DMETH_REC.na, useNA = "always")

# cross tab
table(dmeth_rec = births20to24_step6.pt2$DMETH_REC,
      dmeth_rec.na = births20to24_step6.pt2$DMETH_REC.na,
      useNA = "always")
```


## 7. Changing Variable Class Types

Recall that R stores everything as an object, and each object has a class. Commonly used classes are:
1. logical (e.g., `TRUE`, `FALSE`): `as.logical()`
2. integer: `as.integer()`
3. numeric: `as.numeric()`
4. character: `as.character()`
5. factor (e.g., categorical variables): `factor()`


In our data, we will want to change the following variable class types so that R will recognize them in the way we plan to use them in our analysis:
* `MAGER`: character --> integer
* `MRACEHISP.na`: character --> factor
* `DMETH_REC.na`: numeric --> factor
* `OEGest_R3:` numeric --> factor

Remember, in R everything needs to be assigned because it is an object-oriented language. To do this, we will make a new variable for each of the above using the appropriate class function.

Let's start with `OEGest_R3`, and use the `mutate()` function from the `dplyr` package to create a new variable named `OEGest_R3.fact`.

For factors, we can also assign the level names and the reference value.

```{r}
births20to24_step7 <- births20to24_step6.pt2 %>%
                        mutate(OEGest_R3.fact = factor(OEGest_R3,
                                                          levels = c(2, 1), # list referent 1st
                                                          labels = c("Term", "Preterm"))) # order must match the levels

class(births20to24_step7$OEGest_R3.fact)
levels(births20to24_step7$OEGest_R3.fact)

table(births20to24_step7$OEGest_R3.fact, useNA = "always")
```


### Try it on your own!

Create a new object named `births20to24_step7.pt2` from `births20to24_step7` with a new variable for `MAGER`, `MRACEHISP.na`, and `DMETH_REC.na` with a name structure replacing the `.na` and now ending in `.fact` for factor type (i.e. `MRACEHISP.fact`) and `.int` for integer type (i.e. `MAGER.int`). Keep Non-Hispanic White and Vaginal as the reference values.

The two factors variables currently have the following levels:

| **MRACEHISP.na** | **Label**             |
| :---             | :---                  | 
| 1                | Non-Hispanic White    |
| 2                | Non-Hispanic Black    |
| 3                | Non-Hispanic American Indian and Alaska Native    |
| 4                | Non-Hispanic Asian    |
| 5                | Non-Hispanic Native Hawaiian and Other Pacific Islander   |
| 6                | Non-Hispanic Multiple Races   |
| 7                | Hispanic              |
| 8                | Missing               |

| **DMETH_REC.na** | **Label**             |
| :---             | :---                  | 
| 1                | Vaginal               |
| 2                | C-Section             |
| 9                | Missing               |


**Hint:** you can create multiple new variables within a mutate statement by using a comma between them.

    new_object <- old_object %>%
                      mutate(
                          var1 = oldvar1,
                          var2 = oldvar2
                 )

```{r}
births20to24_step7.pt2 <- births20to24_step7 %>%
                           mutate(
                             # MAGER as integer variable
                             MAGER.int = as.integer(MAGER),
                             # MRACEHISP.na as factor variable
                             MRACEHISP.fact = factor(
                                            MRACEHISP.na, # original variable
                                            levels = c("1", "2", "3", "4", "5", "6", "7"), # list of values with quotes because character type. Does not include NA!
                                            labels = c(
                                              "Non-Hispanic White", 
                                              "Non-Hispanic Black",
                                              "Non-Hispanic American Indian and Alaska Native",
                                              "Non-Hispanic Asian",
                                              "Non-Hispanic Native Hawaiian and Other Pacific Islander",
                                              "Non-Hispanic Multiple Races",
                                              "Hispanic"
                                            ) #labels matching order of levels
                                          ),
                             # DMETH_REC.na as factor variable
                             DMETH_REC.fact = factor(
                                              DMETH_REC.na, # original variable
                                              levels = c(1, 2),
                                              labels = c("Vaginal", "C-Section")
                             )
                           )

```

```{r}
### CODE CHECKS! ###

# Maternal Age ----
class(births20to24_step7.pt2$MAGER.int)
range(births20to24_step7.pt2$MAGER.int)
## table(births20to24_step7.pt2$MAGER.int, useNA = "always") 

# Race and Ethnicity ----
class(births20to24_step7.pt2$MRACEHISP.fact)
levels(births20to24_step7.pt2$MRACEHISP.fact)
## table(births20to24_step7.pt2$MRACEHISP.fact, useNA = "always")

# Delivery Method ----
class(births20to24_step7.pt2$DMETH_REC.fact)
levels(births20to24_step7.pt2$DMETH_REC.fact)
## table(births20to24_step7.pt2$DMETH_REC.fact, useNA = "always")

```

### Try it on your own!

Let's now remove `births20to24`, `births20to24_step5`, `births20to24_step6`, `births20to24_step6.pt2`, `births20to24_step7` from the environment to clean up our workspace a bit. 

Only `births18to24` and `births20to24_step7.pt2` should remain in the environment.

```{r}
rm(births20to24, births20to24_step5, births20to24_step6, births20to24_step6.pt2, births20to24_step7)
```


## 8. Creating Combined Variables

We've already covered the `ifelse()` equivalent for R when creating new variables with a single condition. When we have multi-condition variable creation, `case_when()` is a better fit.



Any diabetes

| **DIAB.recode**     | **RF_PDIAB & RF_GDIAB**           |
| :---                | :---                              |
| Yes                 | RF_PDIAB = Y or/and RF_GDIAB = Y  |
| No                  | (RF_PDIAB = N and RF_GDIAB = N) OR (RF_PDIAB = N & RF_GDIAB = U) OR (RF_PDIAB = U & RF_GDIAB = N)     |
| Missing             | RF_PDIAB = U and RF_GDIAB = U     |


```{r}
births20to24_step8 <- births20to24_step7.pt2 %>%
                         mutate(
                           DIAB.recode = case_when(
                               RF_PDIAB == "Y" | RF_GDIAB == "Y" ~ "Yes", # condition ~ new variable value
                               RF_PDIAB == "N" & RF_GDIAB == "N" ~ "No",
                               RF_PDIAB == "U" & RF_GDIAB == "N" ~ "No",
                               RF_PDIAB == "N" & RF_GDIAB == "U" ~ "No",
                               RF_PDIAB == "U" & RF_GDIAB == "U" ~ NA_character_,
                               TRUE ~ "check" # if none of the conditions are true, set the value to this
                            ),
                           DIAB.recode.fact = factor(
                             DIAB.recode,
                             levels = c("No", "Yes")
                           )
                         )

### Data checks!
class(births20to24_step8$DIAB.recode.fact)

table(births20to24_step8$DIAB.recode.fact, useNA = "always")

births20to24_step8 %>%
    count(DIAB.recode.fact, RF_PDIAB, RF_GDIAB)

```

### Try it on your own!

Create a new object named `births20to24_step8.pt2` from `births20to24_step8` to make the new recode variable for any hypertension and the 10-year age category variables using `mutate()` and `case_when()`. Don't forget to transform them into factor variables named `HYPE.recode.fact` and `MAGER.recode.fact`!


Any hypertension

| **HYPE.recode**     | **RF_PHYPE & RF_GHYPE & RF_EHYP**                    |
| :---                | :---                              |
| Yes                 | RF_PHYPE = Y or/and RF_GHYPE = Y or/and RF_EHYP = Y|
| No                  | Any combination of U and N with at least one N     |
| Missing             | RF_PHYPE = U and RF_GHYPE = U and EHYPE = U        |

10-year age categories

| **MAGER.recode**    | **MAGER.int**   |
| :---                | :---            |
| Under 15 Years      | 12-14           |
| 15-24 Years         | 15-24           |
| 25-34 Years         | 25-34           | 
| 35-44 Years         | 35-44           |
| 45+ Years           | 45-50           |

**Hint:** In R, when listing sequential numbers you can list as `c(1,2,3)` or `c(1:3)`

```{r}
births20to24_step8.pt2 <- births20to24_step8 %>%
                            mutate(
                              # Any Hypertension
                              HYPE.recode = case_when(
                                RF_PHYPE == "Y" | RF_GHYPE == "Y" | RF_EHYP == "Y" ~ "Yes",
                                RF_PHYPE == "N" & RF_GHYPE %in% c("N", "U") & RF_EHYP %in% c("N", "U") ~ "No",
                                RF_PHYPE %in% c("N", "U") & RF_GHYPE == "N" & RF_EHYP %in% c("N", "U") ~ "No",
                                RF_PHYPE %in% c("N", "U") & RF_GHYPE %in% c("N", "U") & RF_EHYP == "N" ~ "No",
                                RF_PHYPE == "U" & RF_GHYPE == "U" & RF_EHYP == "U" ~ NA_character_,
                                TRUE ~ "check"
                              ),
                              HYPE.recode.fact = factor(
                                HYPE.recode,
                                levels = c("No", "Yes")
                              ),
                              # Age Categories
                              MAGER.recode = case_when(
                                MAGER.int %in% c(12:14) ~ "Under 15 Years",
                                MAGER.int %in% c(15:24) ~ "15-24 Years",
                                MAGER.int %in% c(25:34) ~ "25-34 Years",
                                MAGER.int %in% c(35:44) ~ "35-44 Years",
                                MAGER.int %in% c(45:50) ~ "45+ Years"
                              ),
                              MAGER.recode.fact = factor(
                                MAGER.recode,
                                levels = c("Under 15 Years", "15-24 Years", "25-34 Years",
                                           "35-44 Years", "45+ Years"
                                           )
                              )
                            )

```

```{r}
### Data checks! ###

# Any Hypertension ---
class(births20to24_step8.pt2$HYPE.recode.fact)

table(births20to24_step8.pt2$HYPE.recode.fact, useNA = "always")

births20to24_step8.pt2 %>%
    count(HYPE.recode.fact, RF_PHYPE, RF_GHYPE, RF_EHYP)

# Age Categories ---
class(births20to24_step8.pt2$MAGER.recode.fact)

table(births20to24_step8.pt2$MAGER.recode.fact, useNA = "always")

births20to24_step8.pt2 %>%
    count(MAGER.recode.fact, MAGER.int)

```


## 9. Selecting Columns

Similar to other statistical programs, we can clean up our final analytic dataset by only keeping the variables we will be using moving forward.



```{r}
keep_vars <- c("id", "DOB_YY", "OEGest_R3.fact", "MRACEHISP.fact", "MAGER.recode", "DIAB.recode", "HYPE.recode", "DMETH_REC.fact", "DBWT")

births20to24_step9 <- births20to24_step8.pt2 %>%
                        dplyr::select(all_of(keep_vars)) # You can think of `all_of()` as part of the list (select all of these variables in my list)

# check variable names
names(births20to24_step9)
```

### Try it on your own!

Turns out we don't actually need `DBWT`. Let's create a new object named `births20to24.final` from `births20to24_step9` and drop `DBWT` using the drop variable method: (i.e. using a `-` before the variable list).


```{r}
drop_vars <- c("DBWT")

births20to24.final <- births20to24_step9 %>%
                        dplyr::select(-all_of(drop_vars))

```

## 10. Pulling it all together with pipes

Similar to other statistical programming languages, we do not need to have our code run step by step! We can use R pipes (`%>%) to stack our code all in one chunk.

### Try it on your own! 

Combine steps #5 through #8 using pipes to create a new object named `births20to24.final2` from `births18to24`. Recall that consecutive `mutate()` and `filter()` statements can be combined in the same `mutate()` and `filter()` statements with commas.

```{r}
# Define variables for selection in step 9
keep_vars1 <- c("id", "DOB_YY", "OEGest_R3.fact", "MRACEHISP.fact", "MAGER.recode", "DIAB.recode", "HYPE.recode", "DMETH_REC.fact")

# Create final data frame starting with births18to24
births20to24.final2 <- births18to24 %>%
                            # Step 5: Filter to 2020-2024 and valid preterm birth values ----
                            filter(DOB_YY %in% c("2020", "2021", "2022", "2023", "2024"),
                                   OEGest_R3 != 3
                                   ) %>%
                            # Step 6: Coding missing values  ----
                            mutate(MRACEHISP.na = ifelse(MRACEHISP == 8, NA_character_, MRACEHISP),
                                   DMETH_REC.na = ifelse(DMETH_REC == 9, NA_character_, DMETH_REC),
                            # Step 7: Changing variable class types ----
                                   # Preterm birth
                                   OEGest_R3.fact = factor(OEGest_R3,
                                                          levels = c(2, 1), # list referent 1st
                                                          labels = c("Term", "Preterm")),
                                   # Maternal Age
                                   MAGER.int = as.integer(MAGER),
                                   # Race/Ethnicity
                                   MRACEHISP.fact = factor(
                                                  MRACEHISP.na, # original variable
                                                  levels = c("1", "2", "3", "4", "5", "6", "7"), 
                                                  labels = c(
                                                    "Non-Hispanic White", 
                                                    "Non-Hispanic Black",
                                                    "Non-Hispanic American Indian and Alaska Native",
                                                    "Non-Hispanic Asian",
                                                    "Non-Hispanic Native Hawaiian and Other Pacific Islander",
                                                    "Non-Hispanic Multiple Races",
                                                    "Hispanic"
                                                  ) #labels matching order of levels
                                                ),
                                   # # Delivery method
                                   DMETH_REC.fact = factor(
                                                    DMETH_REC.na, # original variable
                                                    levels = c(1, 2),
                                                    labels = c("Vaginal", "C-Section")
                                                ),
                            # Step 8: Creating combined variables ----
                                    # Diabetes Recode 
                                    DIAB.recode = case_when(
                                       RF_PDIAB == "Y" | RF_GDIAB == "Y" ~ "Yes", # condition ~ new variable value
                                       RF_PDIAB == "N" & RF_GDIAB == "N" ~ "No",
                                       RF_PDIAB == "U" & RF_GDIAB == "N" ~ "No",
                                       RF_PDIAB == "N" & RF_GDIAB == "U" ~ "No",
                                       RF_PDIAB == "U" & RF_GDIAB == "U" ~ NA_character_,
                                       TRUE ~ "check" # if none of the conditions are true, set the value to this
                                    ),
                                     DIAB.recode.fact = factor(
                                       DIAB.recode,
                                       levels = c("No", "Yes")
                                     ),
                                    # Hypertension Recode
                                      HYPE.recode = case_when(
                                          RF_PHYPE == "Y" | RF_GHYPE == "Y" | RF_EHYP == "Y" ~ "Yes",
                                          RF_PHYPE == "N" & RF_GHYPE %in% c("N", "U") & RF_EHYP %in% c("N", "U") ~ "No",
                                          RF_PHYPE %in% c("N", "U") & RF_GHYPE == "N" & RF_EHYP %in% c("N", "U") ~ "No",
                                          RF_PHYPE %in% c("N", "U") & RF_GHYPE %in% c("N", "U") & RF_EHYP == "N" ~ "No",
                                          RF_PHYPE == "U" & RF_GHYPE == "U" & RF_EHYP == "U" ~ NA_character_,
                                          TRUE ~ "check"
                                        ),
                                        HYPE.recode.fact = factor(
                                          HYPE.recode,
                                          levels = c("No", "Yes")
                                        ),
                                      # Age Categories
                                        MAGER.recode = case_when(
                                          MAGER.int %in% c(12:14) ~ "Under 15 Years",
                                          MAGER.int %in% c(15:24) ~ "15-24 Years",
                                          MAGER.int %in% c(25:34) ~ "25-34 Years",
                                          MAGER.int %in% c(35:44) ~ "35-44 Years",
                                          MAGER.int %in% c(45:50) ~ "45+ Years"
                                        ),
                                        MAGER.recode.fact = factor(
                                          MAGER.recode,
                                          levels = c("Under 15 Years", "15-24 Years", "25-34 Years",
                                                     "35-44 Years", "45+ Years"
                                                     )
                                        )
                                ) %>%
                              # Step 9: Select Variables 
                              dplyr::select(all_of(keep_vars1))

```

Note we could have piped even earlier (from joins and row binds!) had we not cleared out our environment. If you're interested, try on your own and reach out if you have any difficulties!

## 11. Save data as cleaned R data file

Congrats on having finished your R analytic data set cleaning! For easier use in the future, let's save our clean analytic dataset as an R data file.

```{r, eval = FALSE}
save(births20to24.final, file = "births20to24_final.RData")

```

Before closing out, click the little broom button in the top right environment tab to empty your environment.

# Part 2. 

kable() test - use this for table formatting: https://rmarkdown.rstudio.com/articles_docx.html

```{r}
knitr::kable(table(births18to24$DOB_YY))
```

