---
title: "U.S. Territory Preterm Births during 2020-2023 [COMPLETE EXAMPLE]"
author: "K Labgold"
date: "`r Sys.Date()`"
output:
  word_document: 
    reference_docx: "mod_reference_template.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load packages
pacman::p_load(here, dplyr, flextable, ggplot2)


# After Part 2, Step #3 of data_prep.rmd, load `births20to24_basicsummary`
load(here("training_data_files", "clean", "births20to24_basicsummary.rdata"))

# After Part 2, Step #4 of data_prep.rmd, load `age_raceeth_rates`
load(here("training_data_files", "clean", "age_raceeth_rates.rdata"))

# After Part 3, Step #5 of data_prep.rmd, load XXXXX
```

## Key Findings

  + The number of live births in US Territories increased from `r year.rate20_24[2, 3]` births in 2020 to `r year.rate20_24[2, 3]` births in 2020.  
  + The preterm birth rate was `r year.rate20_24[6,4]` preterm births per 100 live births in 2024, up from `r year.rate20_24[2,4]` preterm births per 100 live births in 2020.  
  + X age group had the highest preterm birth rate during 2020-2024  
  + Race/ethnicity  
  + Preterm birth was higher among xxx smoking, xxx diabetes, xxx hypertension. 
    

## The number of births increased from 2020 to 2024

The number of live births in US Territories increased from `r year.rate20_24[2, 3]` births in 2020 to `r year.rate20_24[2, 3]` births in 2020.

```{r, echo = F}
# Start by renaming our preterm birth rate variable to `Preterm Birth Rate Per 100 Live Births`
# Show the output from each line

year.rate20_24 %>%
  rename("Preterm Births per 100 Live Births" = "preterm_rate_per100") %>%
  flextable() %>%
  theme_vanilla() %>%               # Applies a clean theme
  bg(i = 1, bg = "#5B3DB2", part = "header") %>% # Background color for header
  color(i = 1, color = "white", part = "header") %>%
  bold(j = ~ `Birth Year`) %>%             # Bold a specific column
  set_caption(caption = "Table 1: Preterm Birth Rates by Year") %>%
  width(j = c("Birth Year", "Preterm Births", "Live Births", "Preterm Births per 100 Live Births"),
        width = c(1.5, 1.2, 1.2, 2)) %>%
  ### TRY ON YOUR OWN!!!
  ## try on your own! center the columns
  align(align = "center", part = "all")
  
## take a minute to play around with some of the flextable options
## or try it on your own - look through the documentation to add XXXXX?
## https://ardata-fr.github.io/flextable-book/static/pdf/cheat_sheet_flextable.pdf

```


## The preterm birth rate increased during 2020 to 2024

TRY IT: {INSERT text description}

The preterm birth rate was `r year.rate20_24[6,4]` preterm births per 100 live births in 2024, up from `r year.rate20_24[2,4]` preterm births per 100 live births in 2020.

The 5-year preterm birth rate for 2020-2024 was `r year.rate20_24[1,6]` preterm births per 100 live births.

```{r fig.width = 7, fig.height = 5, dpi=500, include = F}
# Start with this
year.rate20_24 %>%
  filter(`Birth Year` != "Overall 2020-2024") %>%
  mutate(`Birth Year` = as.numeric(`Birth Year`)) %>% # ggplot2 wants year to be ordered 
  ggplot(aes(x = `Birth Year`, y = `preterm_rate_per100`)) +
          geom_line(size = 1.5, color = "steelblue") +
          geom_point(size = 3, color = "black") +
          labs(
            title = "Figure 1. Preterm Birth Rate per 100 Live Births, U.S. Territories, 2020-2024",
            x = "Year",
            y = "Preterm Birth Rate per 100"
          ) +
          theme_minimal()

# Clean Up: remove X title (year is obvious, add limits on the y axis to bring to scale,
# change dot and line colors to match, bold x and y axis tick numbers)
plot2 <- year.rate20_24 %>%
  filter(`Birth Year` != "Overall 2020-2024") %>%
  mutate(`Birth Year` = as.numeric(`Birth Year`)) %>% # ggplot2 wants year to be ordered 
  ggplot(aes(x = `Birth Year`, y = `preterm_rate_per100`)) +
          geom_line(size = 1.5, color = "black") +
          geom_point(size = 3, color = "black") +
          labs(
            title = "Figure 1. Preterm Birth Rate per 100 Live Births, U.S. Territories, 2020-2024",
            x = NULL,
            y = "Rate per 100 Live Births"
          ) +
          theme_minimal() +
          ylim(0, 15) + # change y-axis limits to range from 0 to 15
          theme(
            axis.text = element_text(face = "bold", size = 12),
            axis.title = element_text(face = "bold", size = 13)
          )

plot2
```

```{r fig.width = 7, fig.height = 5, dpi=500, echo = F, warning = F}
### What if we want to highlight the 2024 point with different color & text?
plot2 +
  geom_point(
    data = year.rate20_24 %>% filter(`Birth Year` == "2024"),
    aes(as.numeric(`Birth Year`), preterm_rate_per100),
    color = "#5B3DB2",
    size = 3
  ) +
  geom_text(
    data = year.rate20_24 %>% filter(`Birth Year` == "2024"),
    aes(as.numeric(`Birth Year`), preterm_rate_per100, label = preterm_rate_per100),
    vjust = -1,
    color = "#5B3DB2",
    fontface = "bold"
  )

```

## {INSERT: Age and race/ethnicity summary}

bar chart: preterm birth rate by age

TRY IT: {INSERT text description}

```{r include = F}
# starting just with one variable (maternal age)
age.raceeth.rates %>%
      filter(Variable == "MAGER.recode") %>%
      ggplot(aes(x = Value, y = preterm_rate_per100, fill = "Variable")) +
            geom_col(position = "dodge") +
            labs(
              x = NULL,
              y = "Rate per 100 Live Births",
              fill = "Maternal Age"
            ) +
            theme_minimal()

# let's make the same adjustments from last time 
## Change color & add bar border
## bold text
age.raceeth.rates %>%
      filter(Variable == "MAGER.recode") %>%
      ggplot(aes(x = Value, y = preterm_rate_per100, fill = "Variable")) +
            geom_col(position = "dodge", color = "Black", linewidth = 0.8) + # adding border to the bars
            labs(
              x = NULL,
              y = "Rate per 100 Live Births",
              fill = "Maternal Age"
            ) +
            theme_minimal() +
            ylim(0, 30) + # change y-axis limits to range from 0 to 30
            theme(
                    axis.text = element_text(face = "bold", size = 10),
                    axis.title = element_text(face = "bold", size = 12)
                  ) +
            scale_fill_manual(values = c("lightpink")) 

# Let's drop the legend, and flip the axis so that we can easily read the age groupings
age.raceeth.rates %>%
      filter(Variable == "MAGER.recode") %>%
      ggplot(aes(x = Value, y = preterm_rate_per100, fill = "Variable")) +
            geom_col(position = "dodge", color = "Black", linewidth = 0.8) + # adding border to the bars
            labs(
              x = NULL,
              y = "Rate per 100 Live Births",
              fill = "Maternal Age"
            ) +
            theme_minimal() +
            ylim(0, 30) + 
            theme(
                    axis.text = element_text(face = "bold", size = 10),
                    axis.title = element_text(face = "bold", size = 12),
                    legend.position = "none" # drop legend
                  ) +
            scale_fill_manual(values = c("mediumpurple")) +
            coord_flip() + # flip axis'
            geom_text(
                aes(label = preterm_rate_per100),
                position = position_dodge(width = 0.8),
                hjust = -0.25,
                size = 4,
                fontface = "bold"
              )
```

```{r include = F}
### now make the facet of the two variables
age.raceeth.rates %>%
      ####--- removed the filter to age here ----####
      ggplot(aes(x = Value, y = preterm_rate_per100, fill = "Variable")) +
            geom_col(position = "dodge", color = "Black", linewidth = 0.8) + # adding border to the bars
            labs(
              x = NULL,
              y = "Rate per 100 Live Births",
              fill = "Maternal Age"
            ) +
            theme_minimal() +
            ylim(0, 30) +
            theme(
                axis.text = element_text(face = "bold", size = 10),
                axis.title = element_text(face = "bold", size = 12),
                legend.position = "none" # drop legend
                  ) +
            scale_fill_manual(values = c("mediumpurple")) +
            coord_flip() + # flip axis'
            geom_text(
                aes(label = preterm_rate_per100),
                position = position_dodge(width = 0.8),
                hjust = -0.25,
                size = 4,
                fontface = "bold"
              ) +
            ####---- Adding the faceting here ----####
            facet_grid(Variable ~ .,
                       scales = "free_y", # removes the unused values
                       switch = "y",
                       labeller = labeller(
                         Variable = c(
                           MAGER.recode = "Maternal Age", 
                           MRACEHISP.fact = "Maternal Race & Ethnicity")
                       )
                       ) +
            theme(
              strip.placement = "outside",
              strip.text = element_text(face = "bold", size = 14)
            )
```

```{r fig.width = 7, fig.height = 7, dpi=500, echo = F, warning = F}
### One more step - need to remove the NAs in race/ethnicity!
age.raceeth.rates %>%
      ####--- removed NAs ----####
      filter(!is.na(Value)) %>%
      ggplot(aes(x = Value, y = preterm_rate_per100, fill = "Variable")) +
            geom_col(position = "dodge", color = "Black", linewidth = 0.8) + # adding border to the bars
            labs(
              x = NULL,
              y = "Rate per 100 Live Births",
              fill = "Maternal Age"
            ) +
            theme_minimal() +
            ylim(0, 30) +
            theme(
                axis.text = element_text(face = "bold", size = 10),
                axis.title = element_text(face = "bold", size = 12),
                legend.position = "none" # drop legend
                  ) +
            scale_fill_manual(values = c("mediumpurple")) +
            coord_flip() + # flip axis'
            geom_text(
                aes(label = preterm_rate_per100),
                position = position_dodge(width = 0.8),
                hjust = -0.25,
                size = 4,
                fontface = "bold"
              ) +
            ####---- Adding the faceting here ----####
            facet_grid(Variable ~ .,
                       scales = "free_y", # removes the unused values
                       switch = "y",
                       labeller = labeller(
                         Variable = c(
                           MAGER.recode = "Maternal Age", 
                           MRACEHISP.fact = "Maternal Race/Ethnicity")
                       )
                       ) +
            theme(
              strip.placement = "outside",
              strip.text = element_text(face = "bold", size = 14)
            )
```




## {diabetes, hypertension summary}

side by side bar chart: preterm birth rate by smoking, diabetes, hypertension

? any interest in faceting?

## Recommendations

**try making a figure with icons to insert?


## Data Source and Methods

Include link to where data was pulled

For more information, see documentation available on our [GitHub page](https://github.com/Katie-Labgold).