---
title: "The R-oad Ahead"
format: revealjs
editor: visual
incremental: true
---

## 0. Wrap Up

## 1. Continuing your R Journey

- Regression Models

- Complex Survey Data

- Spatial Analysis

- Interactive Data Dashboards

- Other uses?

## 2. Regression Models

R has a standard formula syntax:

```{r}
#| echo: true
#| eval: false

y ~ x1 + x2 + x1*x2 + log(x)
```

## Key Differences

- No class statement - categorical variables must be coded as factors!

- Model output is an object (classic R!)

- You extract results with functions

## Logistic Regression Example

show model, show output, extract output

## 3. Complex Survey Data

In other stats programs, the survey design is baked into the procedure.

In R, we have to create a survey design object first, then pass that to any summary/modeling functions.

```{r}
#| echo: true 
#| eval: false  

library(survey) # Classic survey package for R  

new.survey.object <- svydesign(   
                        ids = ~psu, #ids - primary sampling unit variable 
                        strata = ~strata, # strata  
                        weights = ~weight, # sampling weight   
                        nest = TRUE # true if PSUs are nested within strata
                      )
```

Commonly used functions in the `survey` package:

- `svymean()`

- `svytable()`

- `svyciprop()`

- `svyglm()` :

- `svycontrast()` :

- `subset()`

## 4. Spatial Analysis

R is a phenomenal tool for mapping and spatial analysis!

- Data manipulation: read/write ESRI shapefiles, prepare R spatial objects (`sf`, `sp`), geocoding[^1]

- Data visualization: choropleth, points, density

- Spatial statistics: hotspots, point pattern analysis, modeling, bayesian methods

[^1]: requires local configuration to protect PHI

::: fragment
**Key packages to get started**[^2]: `sf`, `tmap`, `ggplot2`
:::

[^2]:
    ::: fragment
    Columbia University has a nice summary of [available tools](https://guides.library.columbia.edu/geotools/R)
    :::

## 5. Interactive Data Dashboards

You can use R and [R Shiny](https://rstudio.github.io/shinydashboard/) to develop custom interactive data dashboards that have a complete workflow within the R universe.

<br>

Real-World Application: [Virgin Islands Epidemiology Data Dashboard](https://usvi-doh.shinyapps.io/EpiDashboard/)

## 6. Generally useful resources

## 7. Stay in touch!
