---
title: "Applied R for MCH Epidemiology"
subtitle: "Participant Workbook"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

---

# How to use this workbook

This document is yours to keep. Throughout the day you will fill in code chunks
marked with `## YOUR CODE HERE`. Some chunks are fully scaffolded — you supply
arguments. Others give you a comment describing what to produce and you write
the logic. A few are bonus tasks with no scaffold.

At the end of the day you will have a complete, working, annotated R document
covering the core MCH data tasks we covered together.

**Tip:** Run each chunk as you complete it. Don't wait until the end of a module.

---

# Module 0 | Setup

## Load packages

Run this chunk as-is. If anything fails to load, raise your hand now.

```{r load-packages}
# Core packages for today
library(tidyverse)   # data wrangling and visualization
library(survey)      # complex survey analysis (Module 6)
library(gtsummary)   # formatted tables
library(here)        # file path management

# SAS equivalent: no direct equivalent — SAS uses PROCs rather than libraries
# Stata equivalent: no direct equivalent — packages installed via -ssc install-
# SPSS equivalent: no direct equivalent — SPSS uses extensions/add-ons
```

## Confirm your project structure

```{r check-structure}
# Confirm working directory is set to your R Project root
# If this path looks wrong, stop and let the instructor know
here()

# List files in your data folder
# [INSTRUCTOR: adjust folder name to match distributed project structure]
list.files(here("data"))
```

---

# Module 1 | R Basics

## 1.1 Objects and assignment

```{r objects}
# R uses <- for assignment (= also works but <- is convention)
# SAS equivalent: x = 5; (in data step)
# Stata equivalent: scalar x = 5
# SPSS equivalent: COMPUTE x = 5.

my_number <- 42
my_text   <- "Philadelphia"
my_logical <- TRUE

# Print each object by running its name
my_number
my_text
my_logical
```

## 1.2 Vectors

```{r vectors}
# c() combines values into a vector — the basic building block in R
# SAS equivalent: arrays
# Stata equivalent: no direct scalar equivalent; closest is a variable in a dataset
# SPSS equivalent: no direct equivalent

# --- TASK 1A ---
# Create a numeric vector called birth_counts containing 5 plausible annual
# birth counts for a small state (make up the numbers)
birth_counts <- c(## YOUR CODE HERE)

# Now compute the mean and sum of birth_counts using mean() and sum()
mean(## YOUR CODE HERE)
sum(## YOUR CODE HERE)
```

```{r vector-types}
# Vectors must be a single type — character, numeric, or logical
# Mixing types forces coercion (R will convert silently)

mixed <- c(1, 2, "three")  # what type does this become?
class(mixed)                # check the type

# Indexing with []
# SAS equivalent: array element access e.g. myarray{2}
# Stata equivalent: variable[_n] for row indexing
birth_counts[1]    # first element
birth_counts[2:4]  # elements 2 through 4
```

## 1.3 Data frames

```{r data-frame-basics}
# A data frame is a rectangular table — rows are observations, columns are variables
# This is conceptually identical to a SAS dataset, Stata dataset, or SPSS data file

# We'll work with a small example before loading the real data
example_df <- data.frame(
  year        = c(2018, 2019, 2020, 2021, 2022),
  births      = c(3200, 3150, 3050, 2980, 3100),
  fetal_deaths = c(18, 15, 20, 17, 16)
)

# --- TASK 1B ---
# Use the $ operator to pull the births column and pipe it into summary()
# Pipe operator: |> (base R) or %>% (tidyverse) — both work today
# SAS equivalent: PROC MEANS DATA=example_df VAR births; RUN;
# Stata equivalent: summarize births
# SPSS equivalent: DESCRIPTIVES VARIABLES=births.

example_df$## YOUR CODE HERE |> summary()
```

```{r data-frame-inspect}
# Key inspection functions — run these on example_df

# --- TASK 1C ---
# These are fully scaffolded — just run them and note what each returns

str(example_df)     # structure: types and preview
                    # SAS: PROC CONTENTS; Stata: describe; SPSS: DISPLAY DICTIONARY.

head(example_df)    # first 6 rows
                    # SAS: PROC PRINT DATA=example_df(OBS=6); Stata: list in 1/6

dim(example_df)     # rows, columns
nrow(example_df)    # row count only
ncol(example_df)    # column count only
```

## 1.4 The pipe operator

```{r pipe}
# The pipe passes the output of one function as the first argument to the next
# Think of it as "and then"

# Without pipe:
summary(example_df$births)

# With pipe — reads left to right, same result:
example_df$births |> summary()

# Chaining multiple steps:
example_df$births |> log() |> round(2)

# SAS equivalent: no direct equivalent — SAS uses chained PROCs or macro logic
# Stata equivalent: no direct equivalent — closest is using -quietly- and storing results
# SPSS equivalent: no direct equivalent
```

## 1.5 Functions and help

```{r functions}
# Every operation in R is a function call
# To read documentation: ?function_name or help(function_name)

?mean   # run this in the console — don't include in your final document

# Functions have arguments with defaults
mean(birth_counts)              # uses default na.rm = FALSE
mean(birth_counts, na.rm = TRUE) # explicitly handles missing values
                                  # SAS: MEAN() function ignores missing by default
                                  # Stata: egen mean handles missing by default
                                  # SPSS: MEAN() ignores missing by default
```

---

# Module 2 | Tidyverse and Basic Cleaning

## 2.1 Tidy data

```{r tidy-concept, eval=FALSE}
# Tidy data has:
#   - One row per observation
#   - One column per variable
#   - One value per cell
#
# This is the expected input format for all tidyverse functions
# SAS: analogous to a "vertical" or "tall" dataset structure
# Stata: analogous to data in "long" format
# SPSS: analogous to cases-by-variables structure (standard SPSS format)
```

## 2.2 Load the natality data

```{r load-data}
# [INSTRUCTOR: replace filename with actual distributed dataset]
natality <- read_csv(here("data", "PLACEHOLDER_natality.csv"))

# Quick look
glimpse(natality)
# SAS equivalent: PROC CONTENTS; PROC PRINT DATA=natality(OBS=5); RUN;
# Stata equivalent: describe; list in 1/5
# SPSS equivalent: DISPLAY DICTIONARY. EXECUTE.
```

## 2.3 Core tidyverse verbs

```{r select}
# select() — keep or drop columns
# SAS equivalent: KEEP / DROP statement in data step
# Stata equivalent: keep varlist / drop varlist
# SPSS equivalent: SELECT IF / DELETE VARIABLES

natality_slim <- natality |>
  select(## YOUR CODE HERE)  # [INSTRUCTOR: list the variables to keep]

glimpse(natality_slim)
```

```{r filter}
# filter() — keep rows meeting a condition
# SAS equivalent: WHERE statement or IF in data step
# Stata equivalent: keep if / drop if
# SPSS equivalent: SELECT IF (condition).

# --- TASK 2A ---
# Filter natality_slim to a single state or year
# [INSTRUCTOR: specify which variable and value to filter on]
natality_filtered <- natality_slim |>
  filter(## YOUR CODE HERE == ## YOUR CODE HERE)

nrow(natality_filtered)
```

```{r mutate-recode}
# mutate() creates new variables or modifies existing ones
# case_when() is a vectorized if-else — use for recoding
# SAS equivalent: data step with IF/THEN/ELSE or SELECT/WHEN
# Stata equivalent: generate + replace, or recode command
# SPSS equivalent: COMPUTE + DO IF, or RECODE

# --- TASK 2B ---
# Recode gestational age into a clinical category variable
# [INSTRUCTOR: replace GEST_AGE_VAR with actual variable name and confirm cutpoints]
natality_clean <- natality_slim |>
  mutate(
    gest_cat = case_when(
      ## YOUR CODE HERE < 28                          ~ "Extremely preterm",
      ## YOUR CODE HERE >= 28 & ## YOUR CODE HERE < 32 ~ "Very preterm",
      ## YOUR CODE HERE >= 32 & ## YOUR CODE HERE < 37 ~ "Moderate/late preterm",
      ## YOUR CODE HERE >= 37                          ~ "Term",
      TRUE                                             ~ NA_character_
    )
  )

count(natality_clean, gest_cat)
```

```{r mutate-rate}
# --- TASK 2C ---
# Create a preterm indicator variable (1 = preterm, 0 = term)
# Then compute a simple column for later rate calculation
# [INSTRUCTOR: confirm variable name for gestational age]

natality_clean <- natality_clean |>
  mutate(
    preterm = if_else(## YOUR CODE HERE < 37, 1L, 0L)
    # SAS: preterm = (gest_age < 37);
    # Stata: generate preterm = (GEST_AGE_VAR < 37)
    # SPSS: COMPUTE preterm = (GEST_AGE_VAR < 37).
  )
```

## 2.4 Checking missingness

```{r missing}
# Base R approach — counts NAs per column
colSums(is.na(natality_clean))

# Tidyverse approach
natality_clean |>
  summarise(across(everything(), ~ sum(is.na(.x))))

# SAS equivalent: PROC MEANS NMISS; or PROC FREQ with missing option
# Stata equivalent: misstable summarize
# SPSS equivalent: FREQUENCIES VARIABLES=ALL /MISSING=INCLUDE.
```

---

# Module 3 | Summarization

## 3.1 Counts

```{r counts-basic}
# count() is the fastest path for frequency tables
# SAS equivalent: PROC FREQ DATA=natality_clean; TABLES year; RUN;
# Stata equivalent: tabulate year
# SPSS equivalent: FREQUENCIES VARIABLES=year.

# --- TASK 3A ---
# Count births by year
natality_clean |>
  count(## YOUR CODE HERE)
```

```{r counts-grouped}
# group_by() + summarise() gives more control
# SAS equivalent: PROC MEANS with CLASS statement; or PROC FREQ with multiple variables
# Stata equivalent: bysort year: summarize
# SPSS equivalent: MEANS TABLES=births BY year.

natality_clean |>
  group_by(year) |>
  summarise(n = n()) |>
  ungroup()
# Note: always ungroup() after summarise — leaving groups active causes silent bugs
# No direct equivalent warning in SAS/Stata/SPSS — this is R-specific
```

## 3.2 Prevalence

```{r prevalence}
# --- TASK 3B ---
# Compute count and prevalence of preterm birth stratified by
# [INSTRUCTOR: specify stratification variable, e.g. maternal race/ethnicity]
# [INSTRUCTOR: replace STRAT_VAR with actual variable name]

preterm_prev <- natality_clean |>
  group_by(## YOUR CODE HERE, preterm) |>        # group by stratum and outcome
  summarise(n = n(), .groups = "drop") |>
  group_by(## YOUR CODE HERE) |>                 # re-group by stratum only for proportion
  mutate(
    total = sum(n),
    prop  = n / total,
    pct   = round(prop * 100, 1)
  ) |>
  ungroup() |>
  filter(preterm == 1)                            # keep preterm rows only

preterm_prev

# SAS equivalent:
#   PROC FREQ; TABLES STRAT_VAR * preterm / NOCUM; RUN;
# Stata equivalent:
#   tabulate STRAT_VAR preterm, row
# SPSS equivalent:
#   CROSSTABS TABLES=STRAT_VAR BY preterm /CELLS=ROW COUNT.
```

## 3.3 Rates

```{r rates}
# Rates require linking a numerator dataset to a denominator source
# We'll use a population denominator column here; in Module 4 we join census data

# --- TASK 3C ---
# Compute a crude preterm birth rate per 1,000 live births
# [INSTRUCTOR: confirm whether total births by year/stratum is a column in the data
#  or needs to be computed — adjust scaffold accordingly]

# Step 1: get preterm counts by year
preterm_counts <- natality_clean |>
  group_by(year) |>
  summarise(
    preterm_n = sum(preterm, na.rm = TRUE),
    total_n   = n()
  ) |>
  ungroup()

# Step 2: compute the rate
preterm_counts <- preterm_counts |>
  mutate(
    rate_per_1000 = ## YOUR CODE HERE / ## YOUR CODE HERE * 1000
    # SAS: rate_per_1000 = preterm_n / total_n * 1000;
    # Stata: generate rate_per_1000 = preterm_n / total_n * 1000
    # SPSS: COMPUTE rate_per_1000 = preterm_n / total_n * 1000.
  )

preterm_counts
```

```{r rates-stratified-bonus, eval=FALSE}
# --- BONUS TASK 3D ---
# Stratify the rate by year AND [INSTRUCTOR: specify stratification variable]
# No scaffold — use what you learned above

## YOUR CODE HERE
```

---

# Module 4 | Binding and Joining

## 4.1 Row binding birth and fetal death records

```{r bind-prep}
# Load fetal death data
# [INSTRUCTOR: replace filename]
fetal_deaths <- read_csv(here("data", "PLACEHOLDER_fetal_deaths.csv"))

glimpse(fetal_deaths)
```

```{r bind-rows}
# Before binding, add a record type indicator to each dataset
# This lets you distinguish record sources after the bind
# SAS equivalent: data step SET with an indicator variable before stacking
# Stata equivalent: append using, then generate indicator
# SPSS equivalent: ADD FILES /FILE=* /FILE='fetal.sav'. Then COMPUTE indicator.

# --- TASK 4A ---
# Add record_type variable to each dataset, then bind

natality_typed <- natality_clean |>
  mutate(record_type = "birth")

fetal_typed <- fetal_deaths |>
  mutate(record_type = "fetal_death")

# Now bind — column names must match
# [INSTRUCTOR: if column names differ between datasets, show rename() here first]
combined <- bind_rows(## YOUR CODE HERE, ## YOUR CODE HERE)
```

```{r bind-verify}
# --- TASK 4B ---
# Verify the bind worked — run this chunk as-is
combined |> count(record_type)

# Row counts should sum to original datasets
nrow(natality_typed)
nrow(fetal_typed)
nrow(combined)
```

## 4.2 Joining with census data

```{r load-census}
# Load population denominator data from census
# [INSTRUCTOR: replace filename; confirm key variables: year, state, population]
census <- read_csv(here("data", "PLACEHOLDER_census.csv"))

glimpse(census)
```

```{r left-join}
# left_join() keeps all rows from the left dataset and matches from the right
# Use left_join for almost all vital stats + denominator joins:
#   - you want to keep all birth/death records
#   - not all records may have a census match (flags non-matches rather than dropping)
# SAS equivalent: PROC SQL LEFT JOIN; or data step MERGE with BY variable (be careful)
# Stata equivalent: merge 1:1 year state using census, keep(master match)
# SPSS equivalent: MATCH FILES /FILE=* /TABLE='census.sav' /BY year state.

# --- TASK 4C ---
# Join census population to combined vital stats
# [INSTRUCTOR: specify the key variables, e.g. year + state FIPS]
combined_with_pop <- combined |>
  left_join(census, by = c(## YOUR CODE HERE))
  # e.g. by = c("year" = "year", "state_fips" = "fips")
```

```{r join-verify}
# --- TASK 4D ---
# Verify the join — run as-is, interpret the output

# Row count should be unchanged (left join preserves all left rows)
nrow(combined)
nrow(combined_with_pop)

# anti_join() shows rows that did NOT find a match — investigate any non-zero results
unmatched <- combined |>
  anti_join(census, by = c(## YOUR CODE HERE))  # same by argument as above

nrow(unmatched)
# If nrow(unmatched) > 0, inspect: head(unmatched)
```

---

# Module 5 | Visualization and Tables

## 5.1 ggplot2 basics

```{r ggplot-intro, eval=FALSE}
# ggplot2 uses a grammar of graphics:
#   ggplot(data, aes(x, y, fill, color, ...)) +  # canvas + aesthetic mapping
#   geom_*() +                                    # geometric layer
#   labs() +                                      # labels
#   theme_*()                                     # visual theme

# You build a plot by adding layers with +
# SAS equivalent: PROC SGPLOT — declarative but less composable
# Stata equivalent: twoway — similar layering concept
# SPSS equivalent: GGRAPH or Chart Builder — less flexible
```

## 5.2 Time trend bar chart

```{r bar-chart}
# First prepare the summary data for plotting
# [INSTRUCTOR: adjust variable names to match actual data]
plot_data <- combined_with_pop |>
  filter(record_type == "birth") |>
  group_by(year) |>
  summarise(
    preterm_n = sum(preterm, na.rm = TRUE),
    total_n   = n()
  ) |>
  mutate(rate_per_1000 = preterm_n / total_n * 1000)

# --- TASK 5A ---
# Build a bar chart of preterm birth rate by year
# Scaffold: ggplot() and aes() are started — add geom and labels

ggplot(plot_data, aes(x = year, y = rate_per_1000)) +
  ## YOUR CODE HERE +          # add a bar/column geom: geom_col()
  labs(
    title    = ## YOUR CODE HERE,
    x        = ## YOUR CODE HERE,
    y        = ## YOUR CODE HERE,
    caption  = "[INSTRUCTOR: add data source]"
  ) +
  theme_minimal()
```

```{r chart-modify}
# --- TASK 5B ---
# Modify one visual element of your chart above
# Ideas: change fill color, add value labels, rotate x axis text, change theme
# No scaffold — experiment with one change

## YOUR CODE HERE
```

```{r ggsave}
# --- TASK 5C ---
# Save your figure to the outputs folder
# [INSTRUCTOR: create an outputs/ folder in the distributed project]

ggsave(
  filename = here("outputs", "preterm_trend.png"),
  ## YOUR CODE HERE,   # pass your plot object here, or omit to save last plot
  width  = 8,
  height = 5,
  dpi    = 300
)

# SAS equivalent: ODS GRAPHICS / output delivery system
# Stata equivalent: graph export "filename.png", replace
# SPSS equivalent: OUTPUT EXPORT /CONTENTS EXPORT=VISIBLE /PDF DOCUMENTFILE='file.pdf'.
```

## 5.3 Bonus figure

```{r line-chart-bonus, eval=FALSE}
# --- BONUS TASK 5D ---
# Recreate the time trend as a line chart with points instead of bars
# Hint: geom_line() + geom_point()
# No scaffold

## YOUR CODE HERE
```

```{r figure-reference, eval=FALSE}
# --- REFERENCE: Other geoms worth knowing ---
# geom_histogram()    — distribution of a continuous variable
# geom_boxplot()      — distribution by group
# geom_point()        — scatterplot
# geom_smooth()       — trend line with confidence interval
# facet_wrap(~var)    — small multiples by a grouping variable
# facet_grid(r ~ c)   — two-way small multiples

# Useful extension packages (not installed today — see reference doc):
# patchwork   — combine multiple plots into one figure
# ggridges    — ridge/joy plots for distributions
# ggrepel     — non-overlapping text labels
# scales      — better axis formatting (percent, comma, dollar)
```

## 5.4 Formatted table

```{r table}
# gtsummary produces publication-ready tables with minimal code
# SAS equivalent: PROC TABULATE or PROC REPORT
# Stata equivalent: tabout (user-written) or esttab
# SPSS equivalent: CTABLES

# --- TASK 5E ---
# Produce a formatted frequency table of preterm birth rates by
# [INSTRUCTOR: specify stratification variable]

# [INSTRUCTOR: adjust object name and variables to match actual data]
preterm_prev |>
  select(## YOUR CODE HERE) |>    # select columns you want in the table
  knitr::kable(
    digits  = 1,
    caption = "[INSTRUCTOR: add table title]",
    col.names = c(## YOUR CODE HERE)  # rename columns for display
  )
```

---

# Module 6 | Survey Data Teaser

## Why standard methods are not enough

```{r survey-concept, eval=FALSE}
# The NSCH (National Survey of Children's Health) uses:
#   - Sampling weights     : households selected with unequal probability
#   - Stratification       : sampling within defined geographic/demographic groups
#   - Clustering           : households sampled within clusters, not independently
#
# Ignoring these produces:
#   - Biased point estimates (weights)
#   - Wrong standard errors — usually too small (clustering)
#
# SAS equivalent: PROC SURVEYMEANS, PROC SURVEYFREQ, PROC SURVEYLOGISTIC
# Stata equivalent: svyset then svy: mean / svy: proportion / svy: logistic
# SPSS equivalent: CSPLAN + CSDESCRIPTIVES / CSLOGISTIC
```

## Setting up a survey design object

```{r survey-design, eval=FALSE}
# [INSTRUCTOR: load NSCH data here; provide the actual variable names for
#  weights, strata, and PSU (primary sampling unit)]
# [INSTRUCTOR: this chunk is pre-built — participants run it, do not fill in]

nsch <- read_csv(here("data", "PLACEHOLDER_nsch.csv"))

nsch_design <- svydesign(
  ids     = ~PLACEHOLDER_PSU_VAR,      # cluster/PSU variable
  strata  = ~PLACEHOLDER_STRAT_VAR,    # stratification variable
  weights = ~PLACEHOLDER_WEIGHT_VAR,   # sampling weight
  data    = nsch,
  nest    = TRUE
)

summary(nsch_design)
```

## One weighted estimate

```{r svymean}
# --- TASK 6A (fully scaffolded — just run it) ---
# Compute a weighted prevalence estimate and compare to unweighted

# Unweighted (what you'd get with standard summarise — WRONG for NSCH)
nsch |>
  summarise(
    unweighted_prev = mean(PLACEHOLDER_OUTCOME_VAR, na.rm = TRUE)
  )

# Weighted (correct for complex survey)
svymean(~PLACEHOLDER_OUTCOME_VAR, design = nsch_design, na.rm = TRUE)

# Note the difference in the estimate AND the standard error
```

## Reference: regression with survey data

```{r svyglm-reference, eval=FALSE}
# This is reference code — not a task today
# svyglm() fits regression models accounting for survey design
# SAS equivalent: PROC SURVEYLOGISTIC / PROC SURVEYREG
# Stata equivalent: svy: logistic / svy: regress
# SPSS equivalent: CSLOGISTIC / CSGLM

model <- svyglm(
  PLACEHOLDER_OUTCOME_VAR ~ PLACEHOLDER_PREDICTOR + PLACEHOLDER_COVARIATE,
  design = nsch_design,
  family = quasibinomial()   # use for binary outcomes with survey data
)

summary(model)

# Exponentiated coefficients (odds ratios):
exp(coef(model))
exp(confint(model))
```

---

# Session Info

Run this at the end of the day — useful for reproducibility and troubleshooting.

```{r session-info}
sessionInfo()
```

---

*[INSTRUCTOR: add course contact info, reference doc link, and any follow-up resources here]*
